---
title: "Constrained LS"
author: "Dom Owens"
date: "16/10/2019"
output: html_document
---

Let's demonstrate uses of the KKT conditions for solving optimisation problems.
We consider the constrained least squares problem:
$$min ||\mathbf{Ax - b}||^2 \\ s.t. ||\mathbf{x}||^2 \leq \alpha $$
Solve Lagrangaian $$ L(\mathbf{x}, \lambda) =  ||\mathbf{Ax - b}||^2 + \lambda(||\mathbf{x}||^2-\alpha) $$

If the 2-norm of the least squares solution is less than or equal to alpha, the least squares solution is a minimum; else the solution is the ridge solution $(A^TA + \lambda I)^{-1} A^T \mathbf{b}$.

We require the following bisection function to find lambda: 
```{r}
#INPUT
#f . . . . . . . . . . . . . . . . . . . a scalar function
#lb . . . . . . . . . . . . . ..... the initial lower bound
#ub . . . . . . . . . . . . . . . . . . the initial upper bound
#eps . . . . . . . . . . . . . . . . . tolerance parameter
#OUTPUT
#z . . . . . . . . . . . . . . . . . . a root of the equation f(x)=O
bisect <- function(f,lb,ub,eps){
  if (f(lb) * f(ub) > 0){return("error: postive product")}
  else{
    iter <- 0
    while (ub-lb > eps) {
      z <- (lb+ub)/2
      iter <- iter + 1
      if(f(lb) *f(z) > 0){lb <- z} else {ub <- z}
      print( c(iter,z) )
    }
  }
}

#run with f = x^2 -2 = 0
fn <- function(x){x^2 -2}
bisect(fn, 1, 2, 1e-4)
```

Now for the CLS problem. Here, $f(0)>0$ so we find a point $u>0$ s.t. $f(u)<0$. We can find this by iteratively doubling $u$ until $f(u) < 0$.

```{r}
#INPUT
#A ................... an mxn matrix
#b ................... an m-length vector
#alpha ............... positive scalar
#OUTPUT
# x_cls ............ an optimal solution of
#min{ I IA*x-bl I: I lxl I "2<=alpha}
x_cls <- function(A,b,alpha){
  n <- dim(A)[2] #isolate width
  x_ls <- solve(A, b) #Ax=b
  if(norm(as.matrix(x_ls))^2 <= alpha){x_cls <- x_ls} else{ #set to ls soln, or solve lagrangian
    fn <- function(lambda){norm( solve(t(A)%*%A + diag(lambda,n,n))%*% t(A)%*%b )^2 - alpha }
    u <-1
    while (fn(u) > 0) {
     u <- 2*u #iterate for u
    }
    lambda <- bisect(fn, 0, u, 1e-7) #find lambda with bisection
    x_cls <- solve(t(A)%*%A + diag(lambda,n,n))%*% t(A)%*%b  #set as  regularised soln
  }
}

```
Run on the example

```{r}
A <- matrix( c(1,2,3,1),2,2)
b <- c(2,3)
x_cls(A,b,0.5)
```

